Online Social Cohesion Reflects Real-World Group Action in Syria during the Arab Spring

### 2020-06-03 manuscript.tex
Specify Chicago citation/reference style

Prior to the advent of social media, people coordinated their protest behavior
using phone tree calling sheets and mimeographed flyers, through in-person
meetings of social organizations and churches, and through word of mouth. They
gathered together and hand-made protest signs that reflected their agendas. As
McAdam found, people were more likely to join movements when they had a personal
connection (McAdam, 1986; McAdam \& Paulsen, 1993): Historically, strong
ties have been strong predictors of involvement in high-risk protest activity.

The cohesiveness and strength of social movements in the pre-Internet era could
be operationalized by the number of people turning out to protest and through
the political events that precipitated and resulted from social gatherings \emph{en
masse}. Cohesive social movements were reflected in their cohesive placards and
posters. For example, the \enquote{I am a man} posters held by the men involved in the
1968 sanitation workers' strike reflected their internal cohesion and unified,
goal-oriented movement (Windsor, 2020).

However, the rise of the Internet and its decentralized online networks has
upended theories about recruitment to---and participation in---social movements.
Strong interpersonal ties have yielded to weak impersonal ties that strongly
predict involvement in social movements (Steinert-Threlkeld, 2017). Research
methodologies for measuring social movements have similarly evolved from
ethnographic interviews to \enquote{big data} analyses of language on social media.

@@

Disentangling the complicated relationship between mass online
mobilization---and social mobilization in person---continues to be relevant,
even as our modes of communication change. The political events that
transpired during and after the Arab Spring are still in the process of being
understood. As Zeitzoff (2017) recognizes, social media is
changing conflict, but the mechanisms and relationships by which it does so are
still being understood.

@@

Massive online social cohesion can have real-world social manifestations. Social
media platforms provide one way to enable this kind of online social cohesion by
offering distributed information-gathering and real-time information
dissemination. Given its accessibility and proficiency as an information
exchange tool, Twitter has provided individuals involved in social movements an
opportunity to complement or even replace real-world social mobilization. For
example, during the 2012 Italian protests for global economic change, the riots
in Rome were mirrored by Twitter activity both on the day of the protests and in
the days following them (Vicari, 2013). Tweeting was determined to be more
effective locally in discussing real-world, real-time events than traditional
mainstream media platforms. As more and more individuals rely on social media to
obtain local news and information, it becomes increasingly important to
understand just how these micro-blogs reflect real-world action.

@@

Along with Facebook, Twitter has been credited as essential to the Arab
Spring---both for protesters (as a means to coordinate with one another and to
communicate with the world) and for the government response (to monitor and
track protestors to crack down on dissent) (Comninos, 2011). The events of
Arab Spring have been thoroughly discussed within the political science
literature (e.g., Lutterbeck, 2013; Koehler, 2017), but to our knowledge, the direct \emph{coupling} between
Syrian Twitter cohesion and international action coordination has not been
investigated. In fact, some have been skeptical about the effect of social media
on collective action at all (Seay, 2014).

@@

\begin{quote}
The important story about the 2011
Arab revolts in Tunisia, Egypt, and Libya is not how the globalization of the
norms of civic engagement shaped the protesters' aspirations. Nor is it about
how activists used technology to share ideas and tactics. Instead, the critical
issue is how and why these ambitions and techniques resonated in their various
local contexts. (Anderson, 2011)
\end{quote}

\noindent Others have similarly expressed hesitation about the role of online
organizing. Some have argued that real-life strong ties outweigh online weak ties and
that the power of the strong ties through in-person contact is critical
to recruiting people for higher-risk activities
(Gladwell, 2010; McAdam, 1986).

Interestingly, these views may
actually help to make the case for bridging the gap between social media and
social mobilization. Such views suggest the strength of social ties in an online
environment is inextricably linked to the \emph{context} and \emph{ambitions} of the actors.
Moreover, research suggests that weak ties still retain considerable political
power: While strong ties tend to bind smaller groups, weak ties
tend to bind groups across distances (Granovetter, 1973).

Social media generates critical information about events, including whether and
how a demonstration will take place, how many people participate, and how those
in power (e.g., state, military, security apparatus) respond. Similarly, these
media platforms alert us to more global shifts within a conflict, such as when
fighting moves to a new city or region, the number of casualties and wounded,
and the types of weaponry used by parties to the conflict (Zeitzoff, 2017).
Social media is not entirely event-driven, however. In monitoring social media
channels, government forces may respond to the perception that citizens are
gaining strength and sympathy. Governments have been shown to be quite tolerant
of dissent so long as it does not metamorphose into collective action
(King, Pan, \& Roberts, 2013).

Citizens communicate with each other with increased transparency due to the
ubiquitous availability of social media such as low-cost and egalitarian
platforms like Twitter. Citizens are able to access these tools via mobile
phones and computers with Internet connectivity, providing a stream of real-time
shared information. Citizens can share the location of events like protests and
military incursions, the degree of state repression, and the number of
casualties. In turn, the government---interested in remaining in
power---monitors social media platforms to ascertain emerging threats
(Comninos, 2011), such as where new groups are forming and where existing
groups are strengthening, where gatherings are held, and who the idea
entrepreneurs are.

@@

At the same time, the insulation of online social platforms allows researchers
to detail real-world social dynamics at unprecedented resolution and scale
(Paxton \& Griffiths, 2017). The digital traces left by individuals and collectively as
groups are regarded as a valuable source of information that can offer powerful
insights into states of social disequilibrium, uprisings, and escalations into
civil wars. By examining these conversations, researchers can study the degree
of cohesion between speakers and the alignment between that cohesion and their
collective political behavior.

@@

People join social and political resistance movements because they feel a sense
of connection to the cause and because they have been sufficiently convinced to
participate. Given that people are generally risk-averse, overcoming inertia
requires a strong connection to the group identity or a strong aversion to an
individual, event, or political injustice (Goldstone, 1980, 2011). Political resistance offers participants the
opportunity to \enquote{lose themselves} and relinquish their individual identity in
favor of identification with a larger group identity, contributing to a cohesive
group ready and mobilized for action (Hoffer, 2011).

Previous research on social movements supports the idea that participants are
likely to join movements when they know and have a personal connection to
someone else already involved (McAdam, 1986). The type of collective
contentious behavior that manifests in society is a function of the available
political opportunity structures. The political opportunity structure are
\enquote{comprised of \emph{(sic)} specific configuration of resources, institutional
arrangements and historical precedents for social mobilization, which facilitate
the development of protest movements in some instances and constrain them in
others} (Kitschelt, 1986). What emerges as strategies and tactics in
contentious politics are, in essence, a function of what the system makes
possible. This includes the number of people and their disposition (i.e.
nonviolent or violent), as well as their internal structure. The more cohesive a
group is, the greater threat they pose to the establishment.

Cohesion is also a familiar concept in the study of civil conflicts, as it
pertains to group affiliation and the ability to overcome collective action
problems. Individuals must act in concert---that is, cohesively---to
accomplish the task of challenging the authority of the state (Eisinger, 1973; Francisco, 1995; Olson, 1965; Tilly, 2004).
Recent work has considered rebel recruitment for participation in
civil conflicts as a function of internal cohesion (Gates, 2002):
Rebel movement success is influenced by the degree of group cohesion, which
itself is a function of geographic distance, ideology, and ethnicity.
Within Syria, multiple rebel groups within Syria exist, sourcing fighters locally
from the disaffected Syrian population and internationally through
fighters from transnational terrorist organizations. There is considerable
intergroup competition for participation in the rebel movement and
punishment for defectors for nonparticipation (Gates, 2002).
Foreign fighters would appear to be amply motivated to fight, given their
decision to relocate from a home country that may be located in the region
or much farther away.

Another factor influencing social cohesion during social movements and civil
wars is ethnic identity and the degree of ethnic conflict associated with the
movement. Greater degrees of ethnic conflict increase the salience of ethnic
identity, leading people to identify more strongly with their own ethnic
identity (Sambanis \& Schulhofer-Wohl, 2009). As ethnic conflict intensifies, groups seek to become
more quintessentially like the typical member of their ethnic group. This can
manifest in a number of ways, including in behavior on social media. However,
this may also include a general identification across ethnic groups in a
membership of an opposition to an oppressive leader.

The civil war literature has further explored the role of social cohesion in
detail. Social homogeneity increases the likelihood of social conflict,
especially when the society is polarized (Collier \& Hoeffler, 2004); this is partly
due to nationalist groups' strong identification with a collective identity
(Cederman \& Girardin, 2007) and their social solidarity (Sambanis \& Schulhofer-Wohl, 2009). These
same processes can be similarly important for rebels (Anyanwu, 2002).
Referencing Gurr's (2000) work on minorities at
risk, Regan and Norton (2005) note that

\begin{quote}
a religious or linguistic minority
might suffer disproportionately in a given society, and this form of
aggrievement can lead to unrest across the social lines that distinguish the
minority group. Therefore, while collective cohesion is a function of targeted
maldistributions, it is individual penury that allows for this collective
mobilization. (Regan \& Norton, 2005)
\end{quote}

@@

Just as real-world social cohesion can be quantified, so can online cohesion.
One kind of cohesion can be found in the similarity in language use on social
media. This implicit or explicit similarity in language use can indicate shared
beliefs, thoughts, and feelings, as has been studied in research on language
style in psycholinguistics and social psychology (Brennan \& Clark, 1996; Gonzales, Hancock, \& Pennebaker, 2010; Tausczik \& Pennebaker, 2010). Although previous work on language style has often
focused on individuals and small groups, this can be scaled up to examine
massive social-level changes (Fusaroli et al., 2015; Michel et al., 2011).

The words that comprise social media posts are the first level of cohesion that
we will consider. In Figure \ref{fig:cohesion-in-tweets}, we see four
sequential tweets, each containing four words. Each sequential tweet contains
some content included in the previous tweets. The linguistic construct of
content word overlap measures the relationship across units of meaning---here,
tweets---where identical and semantically similar words are repeated
(Graesser, McNamara, Louwerse, \& Cai, 2004). These repeated lexical items constitute a theme; on Twitter,
this might be a particular hashtag, a name, a location, or other nominal
category. This localized cohesion is referential; words are repeated and
co-referred across tweets, creating a sense of continuity and cohesiveness of
meaning and message.

Typically, these analyses focus on \emph{content words} (e.g., nouns, adjectives,
verbs), which convey more meaningful semantic information than \emph{function words}
(e.g., articles, conjunctions). While function words can provide important
information about complexity and language structure
(Tausczik \& Pennebaker, 2010), content words have more specialized meaning and
are therefore more useful when researchers want to understand \emph{what} people are
discussing, rather than \emph{how} they are discussing it (Graesser et al., 2004; McNamara, Graesser, McCarthy, \& Cai, 2014).

In estimating cohesion in Twitter posts, tweets that share content words
will yield higher cohesion values, even among speakers with opposing
views on the subject matter. Repetition of terms and concepts is captured by
reiteration, which is a type of lexical cohesion in which a subsequent item
repeats a previous item, or refers to it through synonymous language
(Halliday \& Hasan, 1976). In social media, we can conceive of retweets (i.e., reposts
or forwards of previous messages) as a form of reiteration
that increase message cohesion. Since retweets replicate verbatim some part of
another tweet, these tweets would also have higher cohesion. Hashtags function
similarly and are included in our analysis.

@@

The use of similar language can provide insight into \emph{interlocutor cohesion}, or
the degree to which people are talking about similar things. This can---as noted
earlier---include people who are talking about similar things from radically
perspectives, as well as those who are talking about something from the same
perspective. Previous research from political science has investigated these
ideas (Windsor, 2020): The \enquote{Occupy Wall Street} movement has been held up as an
example of a movement with low interlocutor and linguistic cohesion because of
its wide range of signs and slogans, while the \enquote{I am a man} protest (captured by
Ernest Withers during the 1968 sanitation workers' strike in Memphis, TN) or the
\enquote{Black Lives Matter} placards (following the uprising in Ferguson, MO) have high
lexical and interlocutor cohesion (Edgar \& Johnson, 2018).

@@

Given the importance of lexical and interlocutor cohesion for messaging in
political movements, lexical cohesion may be an important bit of trace data for
researchers interested in leveraging social media data to analyze social
movements. Similar to large-scale studies of sociocultural movements (sometimes
referred to as \emph{culturomics}; e.g., Michel et al., 2011), political
scientists can track the ways in which conversations on social media coalesce or
diffuse around similar content words. In this way, it matters relatively little
whether the interlocutors have precisely the same views: The point is to
understand whether the society-level conversation is converging on similar
language.

@@

Real-world collective action is known to bring about a sense of togetherness,
also known as \enquote{collective effervescence} (Durkheim, 1995). This feeling
of being a part of something larger than oneself leads individuals to act in
ways that would not have on their own. Being overcome with this feeling can lead
to violence (Collins, 2009; Spaaij, 2014). Conversely, increased
exposure to violence can result in positive social cohesion. Gilligan, Pasquale, and Samii (2014)
found that exposure to fatal wartime violence resulted in increased prosocial
motivations, including cooperative behaviors, feelings of trust, and overall
community-level social cohesion.

At the same time, behavioral similarity in certain contexts has been
connected to important psychological and social effects. For example, shared
language is associated with rapport (both in face-to-face and computer-mediated
conversations; Niederhoffer \& Pennebaker, 2002; Riordan, Dale, Kreuz, \& Olney, 2011) and
can even be connected to better performance on shared tasks (as long as
the language is relevant to the task at hand; Fusaroli et al., 2012). Shared
movement is associated with greater group bonding and greater willingness
to contribute to the group (Wiltermuth \& Heath, 2009), and shared identity
is associated with a greater willingness to adopt the group's goals and
motivations (Walton, Cohen, Cwir, \& Spencer, 2012).

We should expect to see increased linguistic cohesion among people participating
in online discussions about an idea, event, individual, or movement. Given that
Syrian protesters are known to have coordinated their activity through Twitter
by discussing their own events (e.g., planned protests, active demonstrations)
and their reactions to events happening around them (e.g., government reactions,
international action or inaction), the cohesion in social media should be
expected to correlate with the social cohesion of the group (Zeitzoff, 2011).

This social cohesion can have real and wide-ranging impacts on a political
movement and the people in it. A socially cohesive group is inherently more
threatening to the existing regime because it reflects collective opposition
momentum and the professionalization of potential alternate leadership. Groups
with high social cohesion are more difficult to divide and conquer, and more
cohesive messages may also garner increased international support. Global
insurgency movements exhibit similar patterns in their social structure
(Bohorquez, Gourley, Dixon, Spagat, \& Johnson, 2009). Each violent group has a stable state. Exogenous events
can create perturbations that push the group to become either more cohesive or
more fragmented. Rather than frenetic, patchwork messages via Twitter, emergent
or intentional cohesive online messaging---using similar language, paralleling
in-person messaging (Windsor, 2020)---are a sign of increased organizational
structure, inciting a government response.

Thus, lexical and interpersonal cohesion are hypothesized to be tied to social
and political cohesion. Shared language in an online environment may lead to the
development or expression of a shared identity (Wiltermuth \& Heath, 2009). This,
in turn, may drive individuals' propensity to engage in similar coordinated
behavior on behalf of the group (Walton et al., 2012; Wiltermuth \& Heath, 2009).
Taken together, these two processes may grow movement participants' weak and
strong ties (Gladwell, 2010; Granovetter, 1973; McAdam, 1986), strengthening the movement overall by expanding and
deepening heterogenous ties. Therefore, in the context of social conflict, the
linguistic cohesion of the online conversation should correspond to the degree
of social cohesion on the ground.

Steinert-Threlkeld (2017) modeled a similar relationship
between social media and social mobilization alongside event data. That
groundbreaking work demonstrated that participants on the fringe of a social
movement contribute substantially to the overall level of protest. In addition
to providing new methods to this growing field of study, we build on this
previous work by demonstrating how sociolinguistic cohesion evolves in
response to negative and positive exogenous events.

Political and social cohesion in the world can be modeled using event data which
encodes who does what to whom on a given day. This could include events such as
the government imposing a blockade or fighting with artillery or tanks on
a civilian group. CAMEO (Conflict and Mediation Event Observations) encode
a set of activities with numerical values, rescaled to reflect a
conflict-cooperation scale (Goldstein, 1992; Schrodt \& Hall, 2006). These event
codes include a range of behaviors, from making public statements and appeals,
to demanding, threatening, protesting, and using unconventional mass violence.

@@

Sociolinguistic cohesion is not a static concept. Just as social movements
experience increased participation and setbacks---indicating dynamic
relationships among participants and between participants and the sociopolitical
environment---sociolinguistic cohesion is also dynamic, reflecting the changing
states of the political sphere. Other scholars have noted the diffusion of
messages whereby they reach a critical mass, whereby \enquote{behaviors, beliefs, and
preferences} spread within a population (Barash, Cameron, \& Macy, 2012). Cohesion
changes over time in response to external events and the internal dynamics of
the social movement. This can occur even if the social interaction occurs entirely
online (Centola \& Macy, 2007).

Related previous work has also explored diffusion of hashtags in viral contagion in
Nigeria, a phenomenon similar to what we identify as sociolinguistic cohesion
(Fink, Schmidt, Barash, Cameron, \& Macy, 2016). The prevalence and rate of particular hashtags were
contributed to the overall sociolinguistic cohesion. The more that people
reuse hashtags or common phrases, the greater the sociolinguistic cohesion when
this phenomenon occurs during socially significant times like protests
(State \& Adamic, 2015).

@@

From a dynamical systems perspective, online social cohesion and in-person
action can be considered two complementary components of a single system,
mutually constrained and constantly interacting and influencing one another
(Mahner \& Bunge, 1997). Investigation of the coupling of real-world
mobilization and social cohesion could provide new insight into the power of
social media during mobilization and the ways in which individuals utilize it to
bring about, or respond to, social change. Increasingly, social media platforms
are being utilized during social mobilization (Koopmans, 2004; McAdam, Tarrow, \& Tilly, 2003), but the relationship between virtual and real-world
mobilization is still largely unknown. The present work aims to understand how
physical and virtual movements coexist and interact from a dynamical systems
perspective.

@@

Given the hypothesized interrelationship between the real-world and online
actions---and the notorious messiness of naturally occurring data---it is
critical to find appropriate analysis tools that can identify the complex
interconnections. Traditional linear analyses have led to many notable
discoveries in the domains of psychology and political science, but time series
data pose problems that are not easily overcome with these methods. Events have
historically been conceptualized as linear chains of smaller, sequential events
with one occurrence leading to the next. Certain kinds of data are relatively
well-suited to these analyses, but messy, complex, variable data from real-world
events can pose challenges, from overpowered samples to violation of underlying
statistical assumptions (Paxton \& Griffiths, 2017).

>>>>

Prior to the advent of social media, people coordinated their protest behavior
using phone tree calling sheets and mimeographed flyers, through in-person
meetings of social organizations and churches, and through word of mouth. They
gathered together and hand-made protest signs that reflected their agendas. As
McAdam found, people were more likely to join movements when they had a personal
connection (McAdam 1986; McAdam and Paulsen 1993): Historically, strong
ties have been strong predictors of involvement in high-risk protest activity.

The cohesiveness and strength of social movements in the pre-Internet era could
be operationalized by the number of people turning out to protest and through
the political events that precipitated and resulted from social gatherings \emph{en
masse}. Cohesive social movements were reflected in their cohesive placards and
posters. For example, the \enquote{I am a man} posters held by the men involved in the
1968 sanitation workers' strike reflected their internal cohesion and unified,
goal-oriented movement (Windsor 2020).

However, the rise of the Internet and its decentralized online networks has
upended theories about recruitment to---and participation in---social movements.
Strong interpersonal ties have yielded to weak impersonal ties that strongly
predict involvement in social movements (Steinert-Threlkeld 2017). Research
methodologies for measuring social movements have similarly evolved from
ethnographic interviews to \enquote{big data} analyses of language on social media.

@@

Disentangling the complicated relationship between mass online
mobilization---and social mobilization in person---continues to be relevant,
even as our modes of communication change. The political events that
transpired during and after the Arab Spring are still in the process of being
understood. As Zeitzoff (2017) recognizes, social media is
changing conflict, but the mechanisms and relationships by which it does so are
still largely unknown.

@@

Massive online social cohesion can have real-world social manifestations. Social
media platforms provide one way to enable this kind of online social cohesion by
offering distributed information-gathering and real-time information
dissemination. Given its accessibility and proficiency as an information
exchange tool, Twitter has provided individuals involved in social movements an
opportunity to complement or even replace real-world social mobilization. For
example, during the 2012 Italian protests for global economic change, the riots
in Rome were mirrored by Twitter activity both on the day of the protests and in
the days following them (Vicari 2013). Tweeting was determined to be more
effective locally in discussing real-world, real-time events than traditional
mainstream media platforms. As more and more individuals rely on social media to
obtain local news and information, it becomes increasingly important to
understand just how these micro-blogs reflect real-world action.

@@

Along with Facebook, Twitter has been credited as essential to the Arab
Spring---both for protesters (as a means to coordinate with one another and to
communicate with the world) and for the government response (to monitor and
track protestors to crack down on dissent) (Comninos 2011). The events of
Arab Spring have been thoroughly discussed within the political science
literature (e.g., Lutterbeck 2013; Koehler 2017), but to our knowledge, the direct \emph{coupling} between
Syrian Twitter cohesion and international action coordination has not been
investigated. In fact, some have been skeptical about the effect of social media
on collective action at all (Seay 2014).

@@

\begin{quote}
The important story about the 2011
Arab revolts in Tunisia, Egypt, and Libya is not how the globalization of the
norms of civic engagement shaped the protesters' aspirations. Nor is it about
how activists used technology to share ideas and tactics. Instead, the critical
issue is how and why these ambitions and techniques resonated in their various
local contexts. (Anderson 2011)
\end{quote}

\noindent Others have similarly expressed hesitation about the role of online
organizing. Some have argued that real-life strong ties outweigh online weak ties and
that the power of the strong ties through in-person contact is critical
to recruiting people for higher-risk activities
(Gladwell 2010; McAdam 1986).

Interestingly, these views may
actually help to make the case for bridging the gap between social media and
social mobilization. Such views suggest the strength of social ties in an online
environment is inextricably linked to the \emph{context} and \emph{ambitions} of the actors.
Moreover, research suggests that weak ties still retain considerable political
power: While strong ties tend to bind smaller groups, weak ties
tend to bind groups across distances (Granovetter 1973).

Social media generates critical information about events, including whether and
how a demonstration will take place, how many people participate, and how those
in power (e.g., state, military, security apparatus) respond. Similarly, these
media platforms alert us to more global shifts within a conflict, such as when
fighting moves to a new city or region, the number of casualties and wounded,
and the types of weaponry used by parties to the conflict (Zeitzoff 2017).
Social media is not entirely event-driven, however. In monitoring social media
channels, government forces may respond to the perception that citizens are
gaining strength and sympathy. Governments have been shown to be quite tolerant
of dissent so long as it does not metamorphose into collective action
(King, Pan, and Roberts 2013).

Citizens communicate with each other with increased transparency due to the
ubiquitous availability of social media such as low-cost and egalitarian
platforms like Twitter. Citizens are able to access these tools via mobile
phones and computers with Internet connectivity, providing a stream of real-time
shared information. Citizens can share the location of events like protests and
military incursions, the degree of state repression, and the number of
casualties. In turn, the government---interested in remaining in
power---monitors social media platforms to ascertain emerging threats
(Comninos 2011), such as where new groups are forming, where existing
groups are strengthening, where gatherings are held, and who the idea
entrepreneurs are.

@@

At the same time, the insulation of online social platforms allows researchers
to detail real-world social dynamics at unprecedented resolution and scale
(Paxton and Griffiths 2017). The digital traces left by individuals and collectively as
groups are regarded as a valuable source of information that can offer powerful
insights into states of social disequilibrium, uprisings, and escalations into
civil wars. By examining these conversations, researchers can study the degree
of cohesion between speakers and the alignment between that cohesion and their
collective political behavior.

@@

People join social and political resistance movements because they feel a sense
of connection to the cause. Given that people are generally risk-averse, overcoming inertia
requires a strong connection to the group identity or a strong aversion to an
individual, event, or political injustice (Goldstone 1980, 2011). Political resistance offers participants the
opportunity to \enquote{lose themselves} and relinquish their individual identity in
favor of identification with a larger group, contributing to a cohesive
group ready and mobilized for action (Hoffer 2011).

Previous research on social movements supports the idea that participants are
more likely to join movements when they know and have a personal connection to
someone else already involved (McAdam 1986). The type of collective
contentious behavior that manifests in society is a function of the available
political opportunity structures. The political opportunity structure are
\enquote{comprised of \emph{(sic)} specific configurations of resources, institutional
arrangements and historical precedents for social mobilization, which facilitate
the development of protest movements in some instances and constrain them in
others} (Kitschelt 1986). What emerges as strategies and tactics in
contentious politics are, in essence, a function of what the system makes
possible. This includes the number of people and their disposition (i.e.
nonviolent or violent), as well as their internal structure. The more cohesive a
group is, the greater threat they pose to the establishment.

Cohesion is also a familiar concept in the study of civil conflicts, as it
pertains to group affiliation and the ability to overcome collective action
problems. Individuals must act in concert---that is, cohesively---to
accomplish the task of challenging the authority of the state (Olson 1965; Eisinger 1973; Tilly 2004; Francisco 1995).
Recent work has considered rebel recruitment for participation in
civil conflicts as a function of internal cohesion (Gates 2002):
Rebel movement success is influenced by the degree of group cohesion, which
itself is a function of geographic distance, ideology, and ethnicity.
Within Syria, multiple rebel groups exist, sourcing fighters locally
from the disaffected Syrian population and internationally through
fighters from transnational terrorist organizations. There is considerable
intergroup competition for participation in the rebel movement and
punishment for defectors for nonparticipation (Gates 2002).
Foreign fighters would appear to be amply motivated to fight, given their
decision to relocate from a home country that may be located in the region
or much farther away.

Another factor influencing social cohesion during social movements and civil
wars is ethnic identity and the degree of ethnic conflict associated with the
movement. Greater degrees of ethnic conflict increase the salience of ethnic
identity, leading people to identify more strongly with their own ethnic
identity (Sambanis and Schulhofer-Wohl 2009). As ethnic conflict intensifies, groups seek to become
more quintessentially like the typical member of their ethnic group. This can
manifest in a number of ways, including in behavior on social media. However,
this may also include a general identification across ethnic groups in a
membership of an opposition to an oppressive leader.

The civil war literature has further explored the role of social cohesion in
detail. Social homogeneity increases the likelihood of social conflict,
especially when the society is polarized (Collier and Hoeffler 2004); this is partly
due to nationalist groups' strong identification with a collective identity
(Cederman and Girardin 2007) and their social solidarity (Sambanis and Schulhofer-Wohl 2009). These
same processes can be similarly important for rebels (Anyanwu 2002).
Referencing Gurr's (2000) work on minorities at
risk, Regan and Norton (2005) note that

\begin{quote}
a religious or linguistic minority
might suffer disproportionately in a given society, and this form of
aggrievement can lead to unrest across the social lines that distinguish the
minority group. Therefore, while collective cohesion is a function of targeted
maldistributions, it is individual penury that allows for this collective
mobilization. (Regan and Norton 2005)
\end{quote}

@@

Just as real-world social cohesion can be quantified, so can online cohesion.
One kind of cohesion is the similarity in language use on social
media. This implicit or explicit similarity can indicate shared
beliefs, thoughts, and feelings, as has been studied in research on language
style in psycholinguistics and social psychology (Gonzales, Hancock, and Pennebaker 2010; Tausczik and Pennebaker 2010; Brennan and Clark 1996). Although previous work on language style has often
focused on individuals and small groups, this can be scaled up to examine
massive social-level changes (Michel et al. 2011; Fusaroli et al. 2015).

The words that comprise social media posts are the first level of cohesion that
we will consider. In Figure \ref{fig:cohesion-in-tweets}, we see four
sequential tweets, each containing four words. Each sequential tweet contains
some content included in the previous tweets. The linguistic construct of
content word overlap measures the relationship across units of meaning---here,
tweets---where identical and semantically similar words are repeated
(Graesser et al. 2004). These repeated lexical items constitute a theme; on Twitter,
this might be a particular hashtag, a name, a location, or other nominal
category. This localized cohesion is referential; words are repeated and
co-referred across tweets, creating a sense of continuity and cohesiveness of
meaning and message.

Typically, these analyses focus on \emph{content words} (e.g., nouns, adjectives,
verbs), which convey more meaningful semantic information than \emph{function words}
(e.g., articles, conjunctions). While function words can provide important
information about complexity and language structure
(Tausczik and Pennebaker 2010), content words have more specialized meaning and
are therefore more useful when researchers want to understand \emph{what} people are
discussing, rather than \emph{how} they are discussing it (Graesser et al. 2004; McNamara et al. 2014).

In estimating cohesion in Twitter posts, tweets that share content words
will yield higher cohesion values, even among speakers with opposing
views on the subject matter. Repetition of terms and concepts is captured by
reiteration, which is a type of lexical cohesion in which a subsequent item
repeats a previous item, or refers to it through synonymous language
(Halliday and Hasan 1976). In social media, we can conceive of retweets (i.e., reposts
or forwards of previous messages) as a form of reiteration
that increase message cohesion. Since retweets replicate verbatim some part of
another tweet, these tweets would also have higher cohesion. Hashtags function
similarly and are included in our analysis.

@@

The use of similar language can provide insight into \emph{interlocutor cohesion}, or
the degree to which people are talking about similar things. This can---as noted
earlier---include people who are talking about similar things from radically
perspectives, as well as those who are talking about something from the same
perspective. Previous research from political science has investigated these
ideas (Windsor 2020): The \enquote{Occupy Wall Street} movement has been held up as an
example of a movement with low interlocutor and linguistic cohesion because of
its wide range of signs and slogans, while the \enquote{I am a man} protest (captured by
Ernest Withers during the 1968 sanitation workers' strike in Memphis, TN) or the
\enquote{Black Lives Matter} placards (following the uprising in Ferguson, MO) have high
lexical and interlocutor cohesion (Edgar and Johnson 2018).

@@

Given the importance of lexical and interlocutor cohesion for messaging in
political movements, lexical cohesion may be an important bit of trace data for
researchers interested in leveraging social media data to analyze social
movements. Similar to large-scale studies of sociocultural movements (sometimes
referred to as \emph{culturomics}; e.g., Michel et al. 2011), political
scientists can track the ways in which conversations on social media coalesce or
diffuse around similar content words. In this way, it matters relatively little
whether the interlocutors have precisely the same views: The point is to
understand whether the society-level conversation is converging on similar
language.

@@

Real-world collective action is known to bring about a sense of togetherness,
also known as \enquote{collective effervescence} (Durkheim 1995). This feeling
of being a part of something larger than oneself leads individuals to act in
ways that would not have on their own. Being overcome with this feeling can lead
to violence (Collins 2009; Spaaij 2014). Conversely, increased
exposure to violence can result in positive social cohesion. Gilligan, Pasquale, and Samii (2014)
found that exposure to fatal wartime violence resulted in increased prosocial
motivations, including cooperative behaviors, feelings of trust, and overall
community-level social cohesion.

At the same time, behavioral similarity in certain contexts has been
connected to important psychological and social effects. For example, shared
language is associated with rapport (both in face-to-face and computer-mediated
conversations; Niederhoffer and Pennebaker 2002; Riordan et al. 2011) and
can even be connected to better performance on shared tasks (as long as
the language is relevant to the task at hand; Fusaroli et al. 2012). Shared
movement is associated with greater group bonding and greater willingness
to contribute to the group (Wiltermuth and Heath 2009), and shared identity
is associated with a greater willingness to adopt the group's goals and
motivations (Walton et al. 2012).

We should expect to see increased linguistic cohesion among people participating
in online discussions about an idea, event, individual, or movement. Given that
Syrian protesters are known to have coordinated their activity through Twitter
by discussing their own events (e.g., planned protests, active demonstrations)
and their reactions to events happening around them (e.g., government reactions,
international action or inaction), the cohesion in social media should be
expected to correlate with the social cohesion of the group (Zeitzoff 2011).

This social cohesion can have real and wide-ranging impacts on a political
movement and the people in it. A socially cohesive group is inherently more
threatening to the existing regime because it reflects collective opposition
momentum and the professionalization of potential alternate leadership. Groups
with high social cohesion are more difficult to divide and conquer, and more
cohesive messages may also garner increased international support. Global
insurgency movements exhibit similar patterns in their social structure
(Bohorquez et al. 2009). Each violent group has a stable state. Exogenous events
can create perturbations that push the group to become either more cohesive or
more fragmented. Rather than frenetic, patchwork messages via Twitter, emergent
or intentional cohesive online messaging---using similar language, paralleling
in-person messaging (Windsor 2020)---are a sign of increased organizational
structure, inciting a government response.

Thus, lexical and interpersonal cohesion are hypothesized to be tied to social
and political cohesion. Shared language in an online environment may lead to the
development or expression of a shared identity (Wiltermuth and Heath 2009). This,
in turn, may drive individuals' propensity to engage in similar coordinated
behavior on behalf of the group (Wiltermuth and Heath 2009; Walton et al. 2012).
Taken together, these two processes may increase movement participants' weak and
strong ties (Granovetter 1973; Gladwell 2010; McAdam 1986), strengthening the movement overall by expanding and
deepening heterogenous ties. Therefore, in the context of social conflict, the
linguistic cohesion of the online conversation should correspond to the degree
of social cohesion on the ground.

Steinert-Threlkeld (2017) modeled a similar relationship
between social media and social mobilization alongside event data. That
groundbreaking work demonstrated that participants on the fringe of a social
movement contribute substantially to the overall level of protest. In addition
to providing new methods to this growing field of study, we build on this
previous work by demonstrating how sociolinguistic cohesion evolves in
tandem with negative and positive exogenous events.

Political and social cohesion in the world can be modeled using event data which
encodes who does what to whom on a given day. This could include events such as
the government imposing a blockade or fighting with artillery or tanks on
a civilian group. CAMEO (Conflict and Mediation Event Observations) encode
a set of activities with numerical values, rescaled to reflect a
conflict-cooperation scale (Schrodt and Hall 2006; Goldstein 1992). These event
codes include a range of behaviors, from making public statements and appeals,
to demanding, threatening, protesting, and using unconventional mass violence.

@@

Sociolinguistic cohesion is not a static concept. Just as social movements
experience increased participation and setbacks---indicating dynamic
relationships among participants and between participants and the sociopolitical
environment---sociolinguistic cohesion is also dynamic, reflecting the changing
states of the political sphere. Other scholars have noted the diffusion of
messages whereby they reach a critical mass, whereby \enquote{behaviors, beliefs, and
preferences} spread within a population (Barash, Cameron, and Macy 2012). Cohesion
changes over time in response to external events and the internal dynamics of
the social movement. This can occur even if the social interaction occurs entirely
online (Centola and Macy 2007).

Related previous work has also explored diffusion of hashtags in viral contagion in
Nigeria, a phenomenon similar to what we identify as sociolinguistic cohesion
(Fink et al. 2016). The prevalence and rate of particular hashtags were
contributed to the overall sociolinguistic cohesion. The more that people
reuse hashtags or common phrases, the greater the sociolinguistic cohesion when
this phenomenon occurs during socially significant times like protests
(State and Adamic 2015).

@@

From a dynamical systems perspective, online social cohesion and in-person
action can be considered two complementary components of a single system,
mutually constrained and constantly interacting and influencing one another
(Mahner and Bunge 1997). Investigation of the coupling of real-world
mobilization and social cohesion could provide new insight into the power of
social media during mobilization and the ways in which individuals utilize it to
bring about, or respond to, social change. Increasingly, social media platforms
are being utilized during social mobilization (McAdam, Tarrow, and Tilly 2003; Koopmans 2004), but the relationship between virtual and real-world
mobilization is still largely unknown. The present work aims to understand how
physical and virtual movements coexist and interact from a dynamical systems
perspective.

@@

Given the hypothesized interrelationship between the real-world and online
actions---and the notorious messiness of naturally occurring data---it is
critical to find appropriate analysis tools that can identify the complex
interconnections. Traditional linear analyses have led to many notable
discoveries in the domains of psychology and political science, but time series
data pose problems that are not easily overcome with these methods. Events have
historically been conceptualized as linear chains of smaller, sequential events
with one occurrence leading to the next. Certain kinds of data are relatively
well-suited to these analyses, but messy, complex, variable data from real-world
events can pose challenges, from overpowered samples to violation of underlying
statistical assumptions (Paxton and Griffiths 2017).
